---
title: "Workshop 02. Visualization"
author: "Kira Roberts"
date: "2026-05-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document contains the code needed for the second workshop in the
MB5370 "Introduction to Programming" module.

This workshop focuses on understanding data visualization through the
ggplot package

#### Workspace:

D:/1Techniques/Programing_intro/Programing_intro.Rproj

```{r, install, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
mpg
```

This is the raw plots of the mpg data from which the rest of this
document will modify and work with.

## Creating a ggplot

Using ggplot2, we will plot a graph that shows the size of the engine on
the x-axis and the fuel efficiency on the y-axis.

```{r plots, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

## Graphing template

In this code, you begin with the funtion **ggplot()**. This is used to
create the coordinate system you can add layers to. The first argument
within it adds the data to the graph. We then can use **geom_point**
adds a layer of points to the plot which creates the scatter plot we see
above. There are many **geom** functions that be layered in a plot.

This is a template that can be used widely: ggplot(data = <DATA>) +
<GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))

Without adding any **geom** functions, what we are left with is an empty
plot. We need aesthetic mappings to create a real visual display of the
data.

```{r template, warning=FALSE, message=FALSE}
ggplot(data = mpg) #This creates a blank data window as no axis information has been given to all the display of the included data
```

### Aesthetic mappings

When creating a plot you need two attributes: a geom and aesthetics.
Below we explore a few of these aesthetics.

```{r aesthetics1, warning=FALSE, message=FALSE}
# Change point colour by class
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))

# Manually change point colour
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), colour = "blue")

# Change point size by class
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))

# Change point transparency by class
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))

# Change point shape by class
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))

# Exercise
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class, colour = displ < 5)) #colour separates by whether "displ" is less than 5 - true or false

```

## Faceting

A common practice in scientific literature is to break a complex plot
into several smaller ones. This is called faceting and allows you to
clearly show subsets of the data. We can do this using **facet_wrap()**
but it should only be used for discreet variables.

```{r facets, warning=FALSE, message=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2) # the ~ dictates which variable you want to subset your data with 

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl) #facets plots based on the combination of two variables

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(. ~ cyl) # you can use a "." if you don't want to facet in the rows or columns dimension

```

### Exercise 1

*Read ?facet_wrap. What does nrow do? What does ncol do? What other
options control the layout of the individual panels?*

nrow and ncol both give the number of each, nrow = number of rows and
ncol = number of columns. Scales control the layout by setting the scale
of each of the panels, fixed is the default, free allows the data to
setthe scale and can also be set as a whole or by dimension

## Fitting simple lines

```{r lines, message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) #displays data as points

ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy)) #this takes the points and instead displays it as a smoothed line 

ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv)) #line type creates separate lines based on the "drv" variable 

ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv)) #by grouping the data you show the data is in fact grouped but separated by discreate variables

ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv), #this changes the colour of the lines by their "drv" and thus also separates them
    show.legend = FALSE) #this hide the colour legend

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
#this plot shows both the point data and the data as a smooth line but the multiple lines can be confusing so instead we show it as below
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) +  #we can still change the asetheics of each visualisation
  geom_smooth()

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE) #the subcompact feature then also allows us to select a subset of the data and only plot that 
```

### Exercise 2

1.  *What geom would you use to draw a line chart? A boxplot? A
    histogram? An area chart?*

    The geom_line plots a line that connects each individual data point
    but geom_smooth creates a line of best fits. To make a boxplot you
    use geom_boxplot, a histogram is geom_bar and an area chart is
    geom_area.

2.  *Run this code in your head and predict what the output will look
    like. Then, run the code in R and check your predictions. Will these
    two graphs look different? Why/why not?*

    My prediction is that both these graphs will look the same as they
    are displaying the same information [data = mpg, mapping = aes(x =
    displ, y = hwy)] but the first one sets the information before
    setting how the data is displayed and the second lot of code sets
    the information within each of the display code.

```{r exercise2, message=FALSE, warning=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

## Transforming and Stats

Using a new dataset "Diamonds" we are going to explore transformations
and data summaries. Our first bar chart shows that more diamonds are
available with high quality cuts than low quality cuts.

```{r diamonds, warning=FALSE, message=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))


ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut)) #geoms and stats can often be used interchangeably
```

### Overriding defaults

Like in many other programs, the defaults are not the only way these
functions can work. In fact using defaults can have big implications in
your results, so you should always make an effort to understand any
"black box" outputs. Below we are overriding the default stat which is a
count, to an **identity** which is the raw value of the variable.

```{r tibbles, warning=FALSE, message=FALSE}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
```

You can also override the default mapping from transformed variables to
aesthetics, i.e. displaying a bar chart of the proportion of your total
diamond dataset rather than a count.

```{r overrides, warning=FALSE, message=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1))
# Warning: `stat(prop)` was deprecated in ggplot2 3.4.0.
```

### Plotting statistical details

You can also be transparent in your data limitations or uncertainty by
showing more information on your transformations using
**stat_summary()**

```{r stat.detail, warning=FALSE, message=FALSE}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median)
```

## Aesthetic Adjustments

One of the best ways to convey information in plots is to use aesthetics
like colour or fill.

```{r aesthetics2, warning=FALSE, message=FALSE}
#colour changes the line work of the graphs 
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut)) 

#fill changes the internal colours of the graph 
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut)) 

#changes in the variable can display the data differently 
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```

In these graphs the stacking was automatically done, using a hidden
position argument. We can use these to customize our plots in three
additional ways: **identity** places each object exactly where it falls
in the context of the graph, **fill** which changes the bar height and
**dodge** which forces ggplot2 to not stack things.

```{r aesthetics3, warning=FALSE, message=FALSE}
#To alter transparency (alpha)
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity") 

#To color the bar outlines with no fill color
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")

#fill makes each of the bars the same size (filling the plot)
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill") 

#dodge separates the stacked variables displaying them beside each other
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")

#this adds a small amount of noise to separate any overlapped data points
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter") 
```

## The Layered Grammar of Graphics

The new template for making a ggplot2, now with position adjustments,
stats, and faceting included.
```
ggplot(data = <DATA>) + <GEOM_FUNCTION>( mapping = aes(<MAPPINGS>), stat
= <STAT>, position = <POSITION> ) + <FACET_FUNCTION>
```