---
title: "Workshop 01. Introduction"
author: "Kira Roberts"
date: "2026-05-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Variables and Assignment
#### Workspace: D:/1Techniques/Programing_intro/Programing_intro.Rproj
This document contains the code needed for the first workshop in the MB5370 "Introduction to Programming" module

This section focuses on understanding how data is stored in R and why this matters
```{r}
age <- 25     # shortcut: "alt" + "-" 
first_name <- 'Bill'

Ex1_sum <- 1 + 25.1 + 20.25
```
 
### Functions
Link to all of R's built in functions: "https://stat.ethz.ch/R-manual/R-patched/library/base/html/00Index.html"

```{r}
years_old <- 25.7 
round(years_old) #rounds up
floor(years_old) #rounds down
```

### Arguments 
```{r}
years_old <- 25.765  
round(years_old, 2) #allows you to specify the rounding to 2 decimal points
?round #goes to help page for the rounding function
args(round) #allows you to query arguments for each function
```

### Exercise 
```{r exercise1}
Kyears_old <- 27
paste("Kira is", Kyears_old, "years old")
```


## Debugging
### Misconceptions 
Variables in programs do not work the same way as they do in spreadsheets
```{r misconceptions}
grade <- 55 
total <- grade + 10 
print(total)
grade <- 90 
print(total) # value of total in a spreadsheet will be 100, but in programming a variable holds the value it was assigned (65)
total <- grade + 10 
print(total)  # executed in the way it was defined
```

what will this code do?
```{r test}
p <- 2 
z <- 5
out <- p * z  # What should the value of out be?
print(out)  # What is the value of out? Is it the same as expected?
```



```{r, Debugging errors}
#out <- p * a
# > Error: object 'a' not found -> its an undefined variable.
out <- p * z

# Testing
x <- 1
is.character(x)
is.numeric (x)
```

### Exercise 
```{r exercise2}
# Given code
#my_quiz <- c("uno",
#             "dos",
#             "tres",
#             "cuatro"   #no comma separating values
#             "cinco")
#print (my_quis) #spelling error, should be my_quiz
#str(my_quiz) #defines the string type
#len(my_quiz) #not a real function, assuming its length 

#fixed code
my_quiz <- c("uno",
             "dos",
             "tres",
             "cuatro",
             "cinco")
print (my_quiz)
str(my_quiz)
length(my_quiz)
```

## Data types
### Exercise 
```{r exercise3}
my_numeric_variable = 4.2
class(my_numeric_variable)

my_character_variable = "word"
class(my_character_variable)

my_logical_variable = TRUE
class(my_logical_variable)
  
my_interger_variable = 124L
class(my_interger_variable)
```

## Data structures
```{r}
y <- c(1, 2, 3) #numerical values in a vector
z <- c("Sarah", "Tracy", "Jon") #character  values in a vector
#NOTE: all the values in a vector are the same type

x <- list(1, "a", TRUE) #data (list) and can consist of different types of infomation
x[[2]] #prints the second item in the list

```

## Data frames and tibbles
```{r}
my_dataframe <- data.frame(no = c(1,2,3), c("Tracey", "John", "Pete"), c(TRUE, FALSE, TRUE))
my_dataframe
str(my_dataframe)

my_dataframe$no = as.factor(my_dataframe$no)
str(my_dataframe)
```

## Packages and Libraries
```{r}
#install.packages("tidyverse")
library(tidyverse)
#install.packages("ggplot2")
library(ggplot2)
#install.packages("ggplot","tidyr", "earthtones","redlistr","dplyr" ,"terra")
```

## Coding best practices ####
### Exercise 
1. Read in the .csv dataset into R
2. Filter data by species 
3. Use ggplot to graph counts

# Execute Workshop 1 R script
```{r}
source("../Module_01.R") #reads all the script code from module 1 (found via the file path) and pastes the results here 
```