---
title: "Workshop 01. The Introduction"
author: "Kira Roberts"
date: "2026-05-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This page contains the code needed for the first workshop in the MB5370 "Introduction to Programming" module

This workshop focuses on understanding the RStudio environment, basic syntax and some of the best practices in coding. This includes building an understanding on how data is stored in R and why this matters, we will do this through writing simple code and exploring bugged code.

#### Workspace: D:/1Techniques/Programing_intro/Programing_intro.Rproj

## Objects, Functions and Arguments

At it's simplest, R (and by default RStudio) can act as a calculator for simple arithmetic but it is also a powerful tool for computing and visualizing all sorts of variables.

### Objects and the Assignment Operator

Programming languages allow you to store things as *objects*, or as they're sometimes called *variables*. Storing information in the environment in this way allows you to easily call it back later. To save data in this way, we use an *assignment operator* which looks like an arrow pointing to the left ( <- ). This takes everything on the right hand side and stores it in the object on the left. This operator uses the shortcut: "alt" + "-".


What we see below is just 4 ways this can be utilized. 
 
```{r}
age <- 25   #assigns the value of 25 to the variable "age"

first_name <- 'Bill'  #assigns the character value "Bill" to the object "first_name"

Ex1_sum <- 1 + 25.1 + 20.25 #saves the final value of the sum

fish_lengths <- c(124, 152, 98, 221, 146) #saves the vector of multiple fish lengths
```

### Functions

One of the more powerful aspects of the programming languages used in R, is that they come equipped with many built-in *functions*. These allow you to perform many different tasks. You can see all of R's built in functions using this link: "<https://stat.ethz.ch/R-manual/R-patched/library/base/html/00Index.html>"

```{r}
years_old <- 25.7 
round(years_old) #rounds up
floor(years_old) #rounds down
```

### Arguments

Each function has added sections of code that lets you specify certain aspects of the function. These are called *arguments*, and give you greater control of your output. These are written within the brackets of a function and separated by commas.

```{r}
years_old <- 25.765  
round(years_old, 2) #allows you to specify the rounding to 2 decimal points
args(round) #allows you to query arguments for each function
```

#### Exercise
Using an object, print your name and age using the **paste** function.
```{r exercise1}
Kyears_old <- 27
paste("Kira is", Kyears_old, "years old")
```

## Debugging

Debugging is one of the biggest jobs for professional coders, often consuming 40-60% of their time, as such it is one of the most important skills to learn in coding. Debugging is the process of finding and fixing bugs (errors) in the code. 

### Misconceptions

Variables in programs do not work the same way as they do in spreadsheets. They can be redefined (grade) or not (total) later in your code, this may cause the results you get to be different from what you expect. We can see this below.   

```{r misconceptions}
grade <- 55 
total <- grade + 10 
print(total)

grade <- 90 
print(total) # value of total in a spreadsheet will be 100, but in programming a variable holds the value it was assigned (65)

total <- grade + 10 
print(total)  # executed in the way it was defined
```
The following code shows values being assigned before being put in an equation and the output being given. It should run.
```{r test}
p <- 2 
z <- 5
out <- p * z  # What should the value of out be?
print(out)  # What is the value of out? Is it the same as expected?
```
But this code, hides an error. We were given the error message "Error: object 'a' not found -> its an undefined variable". We can use the above code to fix it. 
```{r, Debugging errors}
#out <- p * a           #error code, commented out
out <- p * z            #fixed code
```

#### Exercise
You are given some code, but it wont run. Debug it. 
```{r exercise2}
# Given code
#my_quiz <- c("uno",
#             "dos",
#             "tres",
#             "cuatro"   #no comma separating values
#             "cinco")
#print (my_quis) #spelling error, should be my_quiz
#str(my_quiz) #defines the string type
#len(my_quiz) #not a real function, assuming its length 

#fixed code
my_quiz <- c("uno",
             "dos",
             "tres",
             "cuatro",
             "cinco")
print (my_quiz)
str(my_quiz)
length(my_quiz)
```



## Data Types and Structures
### Checking Data types
Sometimes it can be useful to check whether your data has been read in correctly and if it is a number - that R recognizes that. The following shows two functions that can help determine what R sees your vector is. 
```{r, Debugging errors}
x <- 1
is.character(x)
is.numeric (x)
```

As well as allowing you to see specifically if your object is a character or numeric, R allows us to determine if the variable is:
**Numeric (numeric / double)**: Continuous decimals (e.g., pH levels: 8.1, 7.95).  
**Integer (integer)**: Whole numbers (e.g., count of sea turtles spotted: 5L). 
**Character (character)**: Text strings enclosed in quotation marks (e.g., species name: "Acanthaster planci"). 
**Logical (logical)**: Boolean values that are either TRUE or FALSE (e.g., Bleaching observed: TRUE).

#### Exercise
```{r exercise3}
my_numeric_variable = 4.2
class(my_numeric_variable)

my_character_variable = "word"
class(my_character_variable)

my_logical_variable = TRUE
class(my_logical_variable)
  
my_interger_variable = 124L
class(my_interger_variable)
```

### Data structures
A vector is the most common data structure seen in R. It is a collection of elements but they all must be of the same type. Lists are similar but can be made up of different types. In both data structures, you can retrieve individual elements using double square brackets to reference their index.
```{r}
y <- c(1, 2, 3) #numerical values in a vector
z <- c("Sarah", "Tracy", "Jon") #character  values in a vector
#NOTE: all the values in a vector are the same type

x <- list(1, "a", TRUE) #data (list) and can consist of different types of infomation
x[[2]] #prints the second item in the list

```

### Data frames and tibbles
Many of us have used spreadsheets before, these rectangular data files are called *data frames* in R. They store data in the same format as excel spreadsheets with rows and columns. Each column can only hold elements of the same type. They can be imported as **.csv** or **.xlsx** files, or you can build one using lists like seen below. 
```{r}
my_dataframe <- data.frame(no = c(1,2,3), c("Tracey", "John", "Pete"), c(TRUE, FALSE, TRUE))
my_dataframe
str(my_dataframe)
```

R usually makes a guess at the variable type, sometimes it gets it right sometimes not. So it's usually a good habit to check it. If it gets it wrong you can change it. The code below takes the "no" column and changes it from a numeric type to a factor. 
```{r}
my_dataframe$no = as.factor(my_dataframe$no)
str(my_dataframe)
```

## Packages and Libraries
To use R packages you must first download or install them, if your machine has already installed them then you'll need to load them for every session. The following downloads and installs several packages but the installation code has been commented out so that the knitting process will work. 
```{r}
#install.packages("tidyverse")
library(tidyverse)

#install.packages("ggplot2")
library(ggplot2)

#install.packages("ggplot","tidyr", "earthtones","redlistr","dplyr" ,"terra")
```
