---
title: "Workshop 01. Introduction"
author: "Kira Roberts"
date: "2026-05-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

#### Workspace: D:/1Techniques/Programing_intro/Programing_intro.Rproj

This document contains the code needed for the first workshop in the MB5370 "Introduction to Programming" module

This workshop focuses on understanding the RStudio environment, basic syntax and some of the best practices in coding. This includes building an understanding on how data is stored in R and why this matters, we will do this through writing simple code and exploring bugged code.

## Objects, Functions and Arguments

At it's simplest, R (and by default RStudio) can act as a calculator for simple arithmetic but it is also a powerful tool for computing and visualizing all sorts of variables.

### Objects and the Assignment Operator

Programming languages allow you to store things as *objects*, or as they're sometimes called *variables*. Storing information in the environment in this way allows you to easily call it back later. To save data in this way, we use an *assignment operator* which looks like an arrow pointing to the left ( <- ). This takes everything on the right hand side and stores it in the object on the left. This operator uses the shortcut: "alt" + "-".
What we see below is just 4 ways this can be utilized. 
 
```{r}
age <- 25   #assigns the value of 25 to the variable "age"

first_name <- 'Bill'  #assigns the character value "Bill" to the object "first_name"

Ex1_sum <- 1 + 25.1 + 20.25 #saves the final value of the sum

fish_lengths <- c(124, 152, 98, 221, 146) #saves the vector of multiple fish lengths
```

### Functions

One of the more powerful aspects of the programming languages used in R, is that they come equipped with many built-in *functions*. These allow you to perform many different tasks. Link to all of R's built in functions: "<https://stat.ethz.ch/R-manual/R-patched/library/base/html/00Index.html>"

```{r}
years_old <- 25.7 
round(years_old) #rounds up
floor(years_old) #rounds down
```

### Arguments

Each function has added sections of code that lets you specify certain aspects of the function. These are called *arguments*, and give you greater control of your output. These are written within the brackets of a function and separated by commas.

```{r}
years_old <- 25.765  
round(years_old, 2) #allows you to specify the rounding to 2 decimal points
args(round) #allows you to query arguments for each function
```

### Exercise

```{r exercise1}
Kyears_old <- 27
paste("Kira is", Kyears_old, "years old")
```

## Debugging

### Misconceptions

Variables in programs do not work the same way as they do in spreadsheets

```{r misconceptions}
grade <- 55 
total <- grade + 10 
print(total)
grade <- 90 
print(total) # value of total in a spreadsheet will be 100, but in programming a variable holds the value it was assigned (65)
total <- grade + 10 
print(total)  # executed in the way it was defined
```

what will this code do?

```{r test}
p <- 2 
z <- 5
out <- p * z  # What should the value of out be?
print(out)  # What is the value of out? Is it the same as expected?
```

```{r, Debugging errors}
#out <- p * a
# > Error: object 'a' not found -> its an undefined variable.
out <- p * z

# Testing
x <- 1
is.character(x)
is.numeric (x)
```

### Exercise

```{r exercise2}
# Given code
#my_quiz <- c("uno",
#             "dos",
#             "tres",
#             "cuatro"   #no comma separating values
#             "cinco")
#print (my_quis) #spelling error, should be my_quiz
#str(my_quiz) #defines the string type
#len(my_quiz) #not a real function, assuming its length 

#fixed code
my_quiz <- c("uno",
             "dos",
             "tres",
             "cuatro",
             "cinco")
print (my_quiz)
str(my_quiz)
length(my_quiz)
```

## Data types

### Exercise

```{r exercise3}
my_numeric_variable = 4.2
class(my_numeric_variable)

my_character_variable = "word"
class(my_character_variable)

my_logical_variable = TRUE
class(my_logical_variable)
  
my_interger_variable = 124L
class(my_interger_variable)
```

## Data structures

```{r}
y <- c(1, 2, 3) #numerical values in a vector
z <- c("Sarah", "Tracy", "Jon") #character  values in a vector
#NOTE: all the values in a vector are the same type

x <- list(1, "a", TRUE) #data (list) and can consist of different types of infomation
x[[2]] #prints the second item in the list

```

## Data frames and tibbles

```{r}
my_dataframe <- data.frame(no = c(1,2,3), c("Tracey", "John", "Pete"), c(TRUE, FALSE, TRUE))
my_dataframe
str(my_dataframe)

my_dataframe$no = as.factor(my_dataframe$no)
str(my_dataframe)
```

## Packages and Libraries

```{r}
#install.packages("tidyverse")
library(tidyverse)
#install.packages("ggplot2")
library(ggplot2)
#install.packages("ggplot","tidyr", "earthtones","redlistr","dplyr" ,"terra")
```

## Coding best practices

### Exercise

1.  Read in the .csv dataset into R
2.  Filter data by species
3.  Use ggplot to graph counts

# Execute Workshop 1 R script

```{r}
#source("../Module_01.R") #reads all the script code from module 1 (found via the file path) and pastes the results here 
```
